---
title: "BIO-463: CosMx Spatial Transcriptomics — Mouse Brain Analysis"
subtitle: "Assignment II"
author: "AI for RNA Lab, University of Bern"
date: "April 2026"
output:
  html_document:
    theme: paper
    highlight: monochrome
    toc: true
    toc_depth: 2
    number_sections: true
  pdf_document:
    toc: true
    toc_depth: '2'
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(warning = FALSE, message = FALSE)
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(cache = TRUE)
knitr::opts_chunk$set(fig.width = 16, fig.height = 10, fig.align = "center")
rm(list = ls())

# ── Package installation and loading ──────────────────────────────────────────
# Run this block once. If a package fails to install, restart R and try again.

if (!require("BiocManager", quietly = TRUE))
  install.packages("BiocManager")

pkgs_bioc <- c("zellkonverter", "SingleCellExperiment",
               "scuttle", "SingleR", "S4Vectors",
               "GenomicRanges", "IRanges", "Biobase")
for (p in pkgs_bioc) {
  if (!require(p, quietly = TRUE, character.only = TRUE))
    BiocManager::install(p)
  library(p, character.only = TRUE)
}

pkgs_cran <- c("Seurat", "ggplot2", "patchwork", "dplyr", "gprofiler2")
for (p in pkgs_cran) {
  if (!require(p, quietly = TRUE, character.only = TRUE))
    install.packages(p)
  library(p, character.only = TRUE)
}

# Note: leidenAlg can be difficult to install from source on some systems
# (particularly macOS arm64). If installation fails, use algorithm = 1
# (Louvain) in FindClusters() instead — results will be very similar.
if (!require("leidenAlg", quietly = TRUE)) {
  tryCatch(
    install.packages("leidenAlg", type = "binary"),
    error = function(e) message("leidenAlg not available — use algorithm=1 below")
  )
}
if (requireNamespace("leidenAlg", quietly = TRUE)) library(leidenAlg)


# ── Colour palette used throughout ───────────────────────────────────────────
# make_palette(n) returns n visually distinct colours regardless of cluster count.
make_palette <- function(n) {
  base <- c(
    "#4E79A7","#F28E2B","#E15759","#76B7B2","#59A14F","#2ecc71","#708090",
    "#800000","#EDC948","#B07AA1","#FF9DA7","#9C755F","#BAB0AC","#D37295",
    "#ff7f50","#e74c3c","#3498db","#9b59b6","#f1c40f","#2c3e50","#27ae60",
    "#1a5276","#48c9b0","#e67e22","#a569bd","#117a65","#cb4335"
  )
  if (n <= length(base)) return(base[seq_len(n)])
  # fall back to hue-spaced colours if more clusters than base colours
  grDevices::hcl.colors(n, palette = "Dark 3")
}

# call this wherever you need colours, e.g.:
#   scale_colour_manual(values = make_palette(nlevels(GE$seurat_clusters)))
```

---

# Background

In this assignment you will analyse **spatial single-cell transcriptomic data**
generated with the
[CosMx Spatial Molecular Imager](https://brukerspatialbiology.com/products/cosmx-spatial-molecular-imager/ffpe-dataset/cosmx-smi-mouse-brain-ffpe-dataset/)
(NanoString/Bruker) from a mouse brain tissue section (coronal hemisphere).

Spatial transcriptomics enables the simultaneous measurement of gene expression
and physical location of cells within a tissue section. Unlike bulk or
dissociated single-cell RNA-seq, it preserves the spatial context of
transcriptional heterogeneity — allowing us to ask not only *which* cell types
are present, but *where* they reside within the tissue.

The CosMx platform preserves the spatial position of each cell within the
tissue, enabling not only the identification of cell types but also the study of
how they are physically organised within anatomical structures.

[![Figure 1: Schematic of imaging-based spatial transcriptomics approaches
(Xenium, MERFISH, CosMx). Source: Wang et al. (2025)](./cosmx_schematic.png){width="55%"}](https://www.nature.com/articles/s41467-025-64990-y)

---

# Task

Using the spatial transcriptomics dataset, produce the following:

**(1) One (or two) scientific figure(s)** addressing the question:

> *What does the data reveal about the transcriptional diversity of cells in the
> mouse brain, and how are distinct cell populations spatially distributed
> across the tissue?*

Your figure should convey a **single, compelling biological message**. You are
free to choose the analytical method(s) you consider most appropriate
(e.g. dimensionality reduction, clustering, spatial mapping), but your choice
must be **justified** — either explicitly in your legend or implicitly through
the figure design.

**(2) A figure legend** written in the style of a research article: concise,
self-contained, and defining all visual elements (axes, colours, groups, method
used).

**(3) A short results paragraph** (3–4 sentences) describing what you observe,
its biological meaning, and its connection to the context of the study.

---

## Grading summary (Total: /20 · Grade: 1 + score/5)

| Component | Marks |
|:---|:---:|
| 1. Figure — method choice and clarity | /6 |
| 2. Figure legend — paper style | /6 |
| 3. Interpretation and link to study context | /6 |
| 4. Scientific communication | /2 |
| **Total** | **/20** |

Assessment prioritises **clarity**, **justification of analytical choices**, and
**biological interpretation**, rather than technical complexity alone.

---

# The dataset

The dataset uses a targeted CosMx Mouse Neuroscience RNA panel of **~950 genes**
measured across thousands of cells. Because the panel is already focused on
biologically relevant targets, we deliberately **do not perform Highly Variable
Gene (HVG) selection** before dimensionality reduction — filtering genes at this
stage would risk discarding signal from rare cell populations.

CosMx data are generated using an
[imaging-based approach](https://www.nature.com/articles/s41587-022-01483-z)
where RNA molecules are detected *in situ* via sequential fluorescent probe
hybridisation cycles. Each cell retains its spatial $(x, y)$ position in the
tissue throughout.

---

# Provided pre-processing pipeline

## What has already been done

You start from a **pre-normalised Seurat object** (`adata_final_format_r.rds`)
produced by an upstream Python pipeline that has already performed:

- QC: removal of negative probes and cells with zero counts
- Library-size normalisation
- log1p transformation

This workflow follows the protocol described in
[Salas et al. (2025)](https://www.biorxiv.org/content/10.64898/2025.12.07.692889v1).

> **Important:** because the data were normalised in Python, we **skip**
> `NormalizeData()` in R to avoid double-transformation. The line
> `obj[["RNA"]]$data <- obj[["RNA"]]$counts` below simply tells Seurat where
> to find the normalised values for downstream steps such as PCA — it does not
> change the data.

## Load the data

```{r load_data}
# Load the prepared Seurat object
# (counts = log-normalised expression; spatial coordinates in metadata)
obj <- readRDS("./adata_final_format_r.rds")

cat("Cells :", ncol(obj), "\n")
cat("Genes :", nrow(obj), "\n")
cat("Metadata columns:", paste(colnames(obj@meta.data), collapse = ", "), "\n")
```

## Check spatial coordinates

CosMx spatial coordinates (`x_um`, `y_um`, in micrometres) are stored in the
Seurat metadata (`obj@meta.data`).

```{r check_metadata}
head(obj@meta.data[, c("x_um", "y_um", "fov",
                       "total_counts", "n_genes_by_counts", "snr")])

cat("X range:", round(range(obj$x_um, na.rm = TRUE)), "µm\n")
cat("Y range:", round(range(obj$y_um, na.rm = TRUE)), "µm\n")
cat("FOVs   :", length(unique(obj$fov)), "\n")
```

```{r spatial_overview, fig.width=12, fig.height=5}
df <- as.data.frame(obj@meta.data)

p1 <- ggplot(df, aes(x = x_um, y = y_um)) +
  geom_point(size = 0.1, color = "#2C3E50") +
  coord_fixed() + theme_minimal() +
  labs(title = "Cell positions — CosMx mouse brain",
       x = "X (µm)", y = "Y (µm)")

p2 <- ggplot(df, aes(x = x_um, y = y_um, color = n_genes_by_counts)) +
  geom_point(size = 0.2) +
  scale_color_viridis_c(name = "Genes\ndetected") +
  coord_fixed() + theme_minimal() +
  labs(title = "Number of genes detected per cell",
       x = "X (µm)", y = "Y (µm)")

p1 + p2
```

## Inspect the normalised expression distributions

```{r inspect_distributions, fig.width=10, fig.height=4}
par(mfrow = c(1, 2))

hist(df$n_genes_by_counts,
     breaks = 80, col = "steelblue", border = NA,
     xlab = "Genes detected per cell", ylab = "Number of cells",
     main = "Genes per cell (post-filtering)", las = 1)
abline(v = median(df$n_genes_by_counts, na.rm = TRUE),
       col = "firebrick", lty = 2, lwd = 2)
legend("topright",
       legend = paste("Median =", round(median(df$n_genes_by_counts, na.rm = TRUE))),
       col = "firebrick", lty = 2, lwd = 2, bty = "n")

hist(log1p(df$total_counts),
     breaks = 80, col = "seagreen", border = NA,
     xlab = "log(Total counts + 1)", ylab = "Number of cells",
     main = "Total counts per cell (log scale)", las = 1)
```

---

# Dimensionality reduction

## PCA

Because the CosMx panel is already targeted (~950 genes), we run PCA on
**all genes** — no HVG selection is applied.

> **Why scale?** Without scaling, highly expressed genes dominate the top PCs
> simply because they have larger absolute variance. `ScaleData()` centres each
> gene to mean 0 and scales to variance 1, ensuring every gene contributes
> equally to PCA.

```{r scale_and_pca, fig.width=10, fig.height=5}
# Make normalised counts available to Seurat's PCA (no re-normalisation)
obj[["RNA"]]$data <- obj[["RNA"]]$counts

# Scale all genes
GE <- ScaleData(obj, features = rownames(obj))
cat("Scaling done:", nrow(obj), "genes scaled.\n")

# Run PCA on all genes
GE <- RunPCA(GE, features = rownames(GE), seed.use = 42)

# Elbow plot: shows variance explained per PC
# Use this to decide how many PCs to retain (look for the "elbow")
ElbowPlot(GE, ndims = 50) +
  ggtitle("Elbow plot — variance explained per PC") +
  theme_minimal()
```

```{r pca_plot, fig.width=6, fig.height=5}
DimPlot(GE, reduction = "pca", dims = c(1, 2), pt.size = 1, alpha = 0.2) +
  ggtitle("PCA — PC1 vs PC2") + theme_minimal()
```

## UMAP

UMAP is run on the top PCs (not on raw gene expression). It produces a 2D
embedding that preserves local neighbourhood structure and is well-suited to
visualising discrete cell populations.

> **Parameters:**  
> `dims`: number of PCs to use — should match your elbow plot choice.  
> `n.neighbors`: size of the local neighbourhood (typical range: 15–50).  
> `min.dist`: controls how tightly cells are packed (lower = tighter clusters).

```{r run_umap, fig.width=6, fig.height=5}
GE <- RunUMAP(GE,
              dims        = 1:50,   # adjust based on elbow plot
              n.neighbors = 30,
              min.dist    = 0.3,
              metric      = "euclidean",
              seed.use    = 42)

DimPlot(GE, reduction = "umap", pt.size = 0.3, alpha = 0.3) +
  ggtitle("UMAP (before clustering)") + theme_minimal()
```

---

# Guided analysis — tools for your assignment

> The code blocks below are **ready to run**. Your task is to:  
> (1) Choose appropriate parameter values where indicated by `# <-- CHOOSE`.  
> (2) Interpret the outputs biologically.  
> (3) Assemble one or two figures that together answer the assignment question.

## Step 1 — Build a cell neighbourhood graph

Before clustering, Seurat constructs a shared nearest-neighbour (SNN) graph:
each cell is connected to its $k$ closest neighbours in PC space.

```{r find_neighbors}
# dims: use the same number of PCs as in RunUMAP above
# k.param: number of neighbours — 20–30 is typical for this dataset size
GE <- FindNeighbors(GE,
                    dims         = 1:50,    # <-- CHOOSE (match UMAP dims)
                    k.param      = 30,      # <-- CHOOSE (neighbours per cell)
                    annoy.metric = "euclidean")
cat("Neighbourhood graph built.\n")
```

## Step 2 — Identify clusters

`FindClusters()` detects communities in the SNN graph. The key parameter is
`resolution`: higher values produce more, smaller clusters.

> **Which algorithm?**  
> `algorithm = 4` (Leiden) is recommended — it avoids poorly connected
> communities. If `leidenAlg` failed to install on your system, use
> `algorithm = 1` (Louvain) as a fallback.

```{r find_clusters}
GE <- FindClusters(GE,
                   resolution = 0.5,    # <-- CHOOSE (try 0.3, 0.5, 0.8, 1.0)
                   algorithm  = 1)      # 4 = Leiden; use 1 if leidenAlg unavailable

cat("Clusters found:", length(unique(GE$seurat_clusters)), "\n")
table(GE$seurat_clusters)
```

## Step 3 — Visualise clusters in UMAP space

```{r umap_clusters, fig.width=7, fig.height=6}
DimPlot(GE,
        reduction  = "umap",
        group.by   = "seurat_clusters",
        cols       = make_palette(nlevels(GE$seurat_clusters)),
        pt.size    = 0.4,
        alpha      = 0.5,
        label      = TRUE,
        label.size = 4) +
  ggtitle("Transcriptional clusters (UMAP)") +
  theme_minimal()
```

## Step 4 — Project clusters onto the tissue

A key strength of CosMx data is the ability to map cluster assignments back onto
the **original tissue coordinates**, revealing anatomically coherent brain
regions (cortex, hippocampus, striatum, etc.).

```{r spatial_clusters, fig.width=10, fig.height=8}
keep       <- !is.na(GE$x_um) & !is.na(GE$y_um)
GE_spatial <- GE[, keep]

df_spatial <- data.frame(
  x       = GE_spatial$x_um,
  y       = GE_spatial$y_um,
  cluster = GE_spatial$seurat_clusters
)

# All clusters together
ggplot(df_spatial, aes(x = x, y = y, colour = cluster)) +
  geom_point(size = 0.1, alpha = 0.7) +
  scale_colour_manual(values = make_palette(nlevels(factor(df_spatial$cluster)))) +
  coord_fixed() +
  theme_minimal(base_size = 11) +
  theme(legend.position = "right", panel.grid = element_blank()) +
  guides(colour = guide_legend(override.aes = list(size = 3, alpha = 1))) +
  labs(title  = "Spatial distribution of transcriptional clusters",
       x = "X (µm)", y = "Y (µm)", colour = "Cluster")
```

```{r spatial_facet, fig.width=14, fig.height=10}
# Individual cluster footprints — useful for identifying spatially coherent types
ggplot(df_spatial, aes(x = x, y = y, colour = cluster)) +
  geom_point(size = 0.15, alpha = 0.6) +
  scale_colour_manual(values = make_palette(nlevels(factor(df_spatial$cluster)))) +
  facet_wrap(~ cluster, ncol = 4) +
  coord_fixed() +
  theme_minimal(base_size = 9) +
  theme(legend.position = "none", panel.grid = element_blank(),
        strip.text = element_text(face = "bold")) +
  labs(title = "Spatial footprint per cluster", x = "X (µm)", y = "Y (µm)")
```

## Step 5 — Identify marker genes per cluster

To interpret clusters biologically, we identify genes that are specifically
highly expressed in each cluster relative to all other cells.

> `FindAllMarkers()` runs a one-vs-rest differential expression test.  
> `only.pos = TRUE` returns only up-regulated genes (the most useful for
> annotation).  
> `min.pct` and `logfc.threshold` filter out weak signals to speed up computation.  
> ⚠️ This step can take **5–10 minutes** on the full dataset.

```{r find_markers, cache=TRUE, eval=FALSE}
markers <- FindAllMarkers(GE,
                          only.pos       = TRUE,
                          min.pct        = 0.25,
                          logfc.threshold = 0.5,
                          test.use       = "wilcox")

# Top 5 markers per cluster
top5 <- markers %>%
  group_by(cluster) %>%
  slice_max(order_by = avg_log2FC, n = 5)

print(top5)
```

## Step 6 — Automated cell type annotation

We use the **`celldex` Mouse RNA-seq reference** to automatically assign a
cell type label to each cell. `SingleR` computes correlations between each
query cell and reference cell type profiles.

> ⚠️ This step can take **10--20 minutes** on the full dataset.  
> **Why not `ZeiselBrainData()`?** That function has a known incompatibility
> with newer versions of `dbplyr` (`collect(Inf)` is no longer valid).
> `celldex` is the recommended replacement and requires no manual
> normalisation step.  
> `label.main` gives coarse labels (neurons, astrocytes, oligodendrocytes,
> endothelial cells, microglia, etc.); use `label.fine` for finer subtypes.

```{r cell_type_annotation, cache=TRUE, eval=FALSE}
# Install celldex if needed
if (!require("celldex", quietly = TRUE))
  BiocManager::install("celldex")
library(celldex)

# Load the mouse RNA-seq reference (downloads once, cached locally after first run)
ref <- celldex::MouseRNAseqData()
cat("Reference label distribution (label.main):\n")
print(table(ref$label.main))

# Convert Seurat object to SingleCellExperiment (required by SingleR)
GE_sce <- as.SingleCellExperiment(GE)

# Run SingleR: correlate each cell against reference profiles
pred <- SingleR(
  test      = GE_sce,
  ref       = ref,
  labels    = ref$label.main,   # switch to ref$label.fine for finer resolution
  de.method = "wilcox"
)

# Transfer predicted labels back to Seurat object
GE$SingleR_label <- factor(pred$pruned.labels)
cat("\nAnnotation done. Label distribution:\n")
print(table(GE$SingleR_label, useNA = "always"))
```

---

# Your assignment deliverables

Using the analysis above as a starting point, produce the following **in a new
section of this R Markdown document** (add your code and text below this
section):

## Figure(s)

Produce **one or two figures** that together convey a single, compelling
biological message about the transcriptional diversity and spatial organisation
of cells in the mouse brain.

Your figure must be:

- visually clear and easy to interpret
- properly labelled (axes, colour legend, annotations)
- designed with meaningful colour choices
- carrying one clear message

## Figure legend

Write a figure legend in the style of a research article. It must be:

- concise and self-contained (a reader unfamiliar with your analysis should
  understand it from the legend alone)
- inclusive of: data type, method used, what is shown (clusters / cell types /
  genes), and a definition of all visual elements (colours, axes, groups)

## Results paragraph

Write a short results paragraph (3–4 sentences) that:

1. Describes the main observation
2. Explains its biological meaning
3. Connects your finding to the context of the study (mouse brain cell
   organisation)

---

*BIO-463 Genomics & Bioinformatics — Assignment II*
